---
output: github_document
bibliography: inst/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "65%",
  fig.align = "center"
)
```

## The `RcppCensSpatial` package

<!-- badges: start -->
<!-- badges: end -->

The `RcppCensSpatial` package fits a spatial censored linear regression model using the Expectation-Maximization (EM) [@dempster1977maximum], Stochastic Approximation EM (SAEM) [@delyon1999convergence], or Monte Carlo EM (MCEM) [@wei1990monte] algorithm. These algorithms are widely used to compute the maximum likelihood (ML) estimates for incomplete data problems. The EM algorithm computes the ML estimates when a closed expression for the conditional expectation of the complete-data log-likelihood function is available. In the MCEM algorithm, the conditional expectation is substituted by a Monte Carlo approximation based on many independent simulations of the partially observed data. In contrast, the SAEM algorithm splits the E-step into simulation and integration steps.

This package also approximates the standard error of the estimates using the method developed by @louis1982finding and supports missing values on the dependent variable. Moreover, it has a function that performs spatial prediction in new locations. It also allows computing the covariance matrix and the distance matrix. For more information about the model formulation and estimation, please see @ordonez2018geostatistical and @valeriano2021likelihood.

The `RcppCensSpatial` library provides the following functions:

* `CovMat`: computes the spatial covariance matrix.
* `dist2Dmatrix`: computes the Euclidean distance matrix for a set of coordinates.
* `EM.sclm`: fits a spatial censored linear regression model via the EM algorithm.
* `MCEM.sclm`: fits a spatial censored linear regression model via the MCEM algorithm.
* `SAEM.sclm`: fits a spatial censored linear regression model via the SAEM algorithm.
* `predict.sclm`: performs spatial prediction in a set of new locations.
* `rCensSp`: simulates censored spatial data for an established censoring rate.

`print`, `summary`, `predict`, and `plot` functions also work for the `sclm` class.

Next, we will describe how to install the package and use all the previous methods in an artificial example.

### Installation

The released version of `RcppCensSpatial` from [CRAN](https://CRAN.R-project.org) can be installed with:

``` r
install.packages("RcppCensSpatial")
```
### Example
In the following example, we simulate a dataset of length n = 220 from the spatial linear regression model considering three covariates and the exponential correlation function to deal with the variation between spatial points. In order to evaluate the prediction accuracy, the dataset is train-test split. The training data consists of 200 observations, with 5% censored to the left, while the testing data contains 20 observations.

```{r example1}
library(RcppCensSpatial)

set.seed(12341)
n = 220
x = cbind(1, runif(n), rnorm(n))
coords = round(matrix(runif(2*n, 0, 15), n, 2), 5)
dat = rCensSp(beta=c(1,2,-1), sigma2=1, phi=4, nugget=0.50, x=x, coords=coords,
              cens='left', pcens=.05, npred=20, cov.model="exponential")
# Proportion of censoring
table(dat$Data$ci)
```

For comparison purposes, we fit the spatial censored linear model for the simulated data using three approaches: EM, MCEM, and SAEM algorithm. Each method considers the same maximum number of iterations `MaxIter=300`, and the spatial correlation function used in the simulation process, i.e., `type='exponential'`, the default value. Other types of spatial correlation functions available are `'matern'`, `'gaussian'`, and `'pow.exp'`.

```{r example2}
data1 = dat$Data

# EM estimation
fit1 = EM.sclm(data1$y, data1$x, data1$ci, data1$lcl, data1$ucl, data1$coords, 
               phi0=3, nugget0=1, MaxIter=300, error=1e-4)
fit1$tab

# MCEM estimation
fit2 = MCEM.sclm(data1$y, data1$x, data1$ci, data1$lcl, data1$ucl, data1$coords, 
                 phi0=3, nugget0=1, MaxIter=300, nMax=1000, error=1e-4)
fit2$tab

# SAEM estimation
fit3 = SAEM.sclm(data1$y, data1$x, data1$ci, data1$lcl, data1$ucl, data1$coords, 
                 phi0=3, nugget0=1, M=10, error=1e-4)
fit3$tab
```

Note that the estimates obtained for each parameter are similar and close to the true parameter value, except for the first regression coefficient, which was estimated close to 0.70, while the true value was equal to 1. 

Moreover, generic functions `print` and `summary` return some results of the fit for the `sclm` class, such as the estimated parameters, standard errors, the effective range, the information criteria, and some convergence details.

```{r example3}
print(fit3)
```
On the other hand, the function `plot` provides convergence graphics for the parameters.

```{r example4}
plot(fit3)
```

Now, we compute the predicted values for each fitted model for the testing data and compare the mean squared prediction error (MSPE).

```{r example5}
data2 = dat$TestData
pred1 = predict(fit1, data2$coords, data2$x)
pred2 = predict(fit2, data2$coords, data2$x)
pred3 = predict(fit3, data2$coords, data2$x)

# Cross-validation
mean((data2$y - pred1$predValues)^2)
mean((data2$y - pred2$predValues)^2)
mean((data2$y - pred3$predValues)^2)
```

### References
